Reproducing Activity Recognition of Weight Lifting Exercises
============================================================

This exercise for the Coursera [Practical Machine Learning](https://www.coursera.org/course/predmachlearn) course aims to predict types of weight lifting exercises from measurements of sensors.

The original data comes from:

*[Qualitative Activity Recognition of Weight Lifting Exercises](http://groupware.les.inf.puc-rio.br/public/papers/2013.Velloso.QAR-WLE.pdf)* 
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th Augmented Human (AH) International Conference in cooperation with ACM SIGCHI (Augmented Human'13) . Stuttgart, Germany: ACM SIGCHI, 2013.

Data preparation
----------------

First R needs to read the training and test data provided in the correct location.

```{r}
setwd("C:/Users/HB/Weight_Lifting/")
datalocation <- "F:/Documents/Coursera/Practical Machine Learning/"
training <- read.csv(paste0(datalocation, "pml-training.csv"))
testing  <- read.csv(paste0(datalocation, "pml-testing.csv"))
```

Cleaning the training data involves removing any variables which have missing values, any variables which are not-numeric (except for the `classe` variable, which is what is being predicted) and the initial four variables `X`, `raw_timestamp_part_1`, `raw_timestamp_part_2`, and `num_window` which just give an order to the observations and may cause leakage if retained.  See how the dimension of the training dataframe reduces at each step:

```{r}
dim(training) 
trainingnotNA <- training[,colSums(is.na(training))==0]
dim(trainingnotNA) 
trainingnumeric <- trainingnotNA[, sapply(trainingnotNA, is.numeric)]
dim(trainingnumeric) 
trainingtouse <- cbind(trainingnumeric[,-(1:4)], classe=training$classe)
dim(trainingtouse)
```

The following numeric variables have been retained to training and predict `classe`:

```{r}
variablestouse <- names(trainingtouse)[names(trainingtouse) != "classe"]
variablestouse
```

Some R libraries are needed to perform Random Forest machine learning.

```{r}
library(caret)
library(randomForest)
library(e1071)
```

Model fitting 
-------------

The original analysis by Velloso *et al* states it used a Random Forest approach, so this analysis does too. The model fitting uses the `caret` package to apply the Random Forest method to train the machine learning model using what remains of the training data after cleaning. This takes some time and uses a large amount of memory, at least on this machine.      

It also uses 3-fold cross validation to give an estimate of out-of-sample error. Setting the seed allows reproducability.

```{r}
set.seed(2014)
fitControl <- trainControl(method="cv", number=3)
modelFit   <- train(classe ~ ., data=trainingtouse, method="rf", 
                    trControl=fitControl)
```

Results
-------

The results of the model fit suggests that it has a perfect in-model fit on all 19622 observations, producing a confusion matrix where all values are on the diagonal.

The 3-fold cross-validation suggests that out-of-model predictions should be almost perfect too: the only clues that it might not be completely perfect are the small but poositive standard deviations for accuracy and Kappa. Even so, the cross-validation statistics suggest that the should still be more than 99% accurate.  

```{r}
print(modelFit)
sum(predict(modelFit, training) == training$classe)  # in-model correct
sum(predict(modelFit, training) != training$classe)  # in model failures
confusionMatrix(predict(modelFit, training), training$classe)
```

There is also information about the importance of the different variables used for prediction, which can be seen on this plot. Belt measurements tend to be more important and gyro measurements tend to be relatively less important. 

```{r fig.width=9, fig.height=9}
plot(varImp(modelFit, scale=FALSE), xlim=c(0, 1000),
     main="Variable importance of 52 used in fitted model")
```

Prediction of test set
----------------------

A fitted model allows prediction from the test set. The values need to be converted from a factor variable vector to characters before submission.

```{r}
answers <- as.character(predict(modelFit, testing))
answers
```

20 observations are not sufficient to see precisely how close the out-of-model accuracy is in fact to 100% as one error would drop the result to 95%.  In fact, all 20 predictions were subsequently validated as correct. 
